---
title: "DonghanLiuHW02"
author: "Donghan Liu, Donghan2"
date: "2018/2/4"
output:
  html_document:
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_knit$set(echo = TRUE,message = FALSE, warning = FALSE,root.dir = "~/Stat480/RDataScience/AirlineDelays")
```

#Solution

##Exercise 1

```{r}
library(biganalytics)
library(data.table)
library(kableExtra)
library(knitr)
```

```{r}
data0708 = attach.big.matrix("air0708.desc")
may0708 = data0708[data0708[,"Month"] == 5,]
nodelay = c(rep(0,7))
for (i in 1:7){
  nodelay[i] = nrow(may0708[may0708[,'ArrDelay']<=0 & may0708[,'DayOfWeek']==i & may0708[,'DepDelay']<=0 & may0708[,'WeatherDelay']<=0 & may0708[,'Cancelled']<=0
                             & may0708[,'LateAircraftDelay']<=0 & may0708[,'NASDelay']<=0 & may0708[,'SecurityDelay']<=0 & may0708[,'CarrierDelay']<=0 & may0708[,'Diverted']<=0,])
}

dayofweek = c('Mon','Tue', 'Wed', 'Thu','Fri', 'Sat', 'Sun')
nodelaytable = data.frame(NumOfNoDelays = nodelay, DayOfWeek = dayofweek)

kable_styling(kable(nodelaytable, format = "html", digits = 2), full_width = FALSE)


plot(nodelay,type = 'o',main = 'No Any Kind of Delays Regarding Day of Week in 2007&2008',xlab = 'Day of Week',ylab = 'Number of Flight that No Any Kind of Delays')
```

According to the above plot, the Thursday has the most number in no any kind of dalay flights and it potentially would be the best day to travel. 


##Exercise 2

```{r}
noarrdelay = c(rep(0,12))
noarrdelayrate = c(rep(0,12))
for (i in 1:12){
  noarrdelay[i] = nrow(data0708[data0708[,'Month'] == i & data0708[,'ArrDelay'] <= 0,])
  noarrdelayrate[i] = (nrow(data0708[data0708[,'Month'] == i & data0708[,'ArrDelay'] <= 0,])/nrow(data0708[data0708[,'Month'] == i,]))*100
}
monthofyear = c('Jan','Feb', 'Mar', 'Apr','May', 'June', 'July','Aug' ,'Sep','Oct','Nov','Dec')
noarrdelaytable = data.frame(NumOfNoArrDelays = noarrdelay,RateofNoArrDelays_Percent = noarrdelayrate, MonthOfYear = monthofyear)

kable_styling(kable(noarrdelaytable, format = "html", digits = 2), full_width = FALSE)

par(mfrow = c(2, 1))
plot(noarrdelay,type = 'o',main = 'No Arrival Delays Regarding Month of Year in 2007&2008',xlab = 'Month of Year',ylab = 'Number of Flight that No Arrival Delays')
plot(noarrdelayrate,type = 'o',main = 'No Arrival Delays Percentage Regarding Month of Year in 2007&2008',xlab = 'Month of Year',ylab = 'Number of Flight that No Arrival Delays')
```

Regarding the above graph and table, the month of Sep has the highest number and rate of no delays, so other than May, Sep is a good option to travel based on avoiding the arrival delays.


##Exercise 3

```{r}
arrdelay = data0708[,'ArrDelay']
flightime = data0708[,'CRSElapsedTime']
dist = data0708[,'Distance']

#CRSElapsedTime
fitarr = lm(arrdelay ~ flightime)
summary(fitarr)
```
Based on the above summary table, we could see that the estimated paramater coefficient is 0.0047651 (positive relationship), the p-value is less than 2e-16, and the R_squared is 7.352-05, which represent the percentage that CRSElapsedTime could explain the arrival delays in 2007-2008

```{r}
#Distance
fitdist = lm(arrdelay ~ dist)
summary(fitdist)
```
As the above table shows, we could see that the estimated paramater coefficient is 1.453e-04 (week positive relationship), the p-value is 2.67e-15, and the R_squared is  4.423e-06, which represent the percentage that Distance could explain the arrival delays in 2007-2008.

Overall, both of predictors are not the good predictors for the arrival delays, since the R_squared is relatively small. However, if we compare these two predictors, the model that predict arrival delays with CRSElapsedTime is better, since 1. the p-value of the CRSElapsedTime is lower, meaning that it is more significant 2. the model of CRSElapsedTime R_squared is larger, indicating that it explains more to the arrival delays.





