---
title: "DonghanLiuHW04"
author: "Donghan Liu, Donghan2"
date: "2/18/2018"
output:
  html_document:
    theme: readable
    toc: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_knit$set(echo = TRUE,message = FALSE, warning = FALSE,root.dir = "~/Stat480/RDataScience/Chapter3")
```

```{r, message = FALSE, warning = FALSE,echo = FALSE}
# The following setup code is based on the code for Chapter 3 of 
# 
# Nolan, Deborah and Temple Lang, Duncan. Data Science in R: A Case Studies Approach to 
# Computational Reasoning and Problem Solving. CRC Press, 2015. 
# http://rdatasciencecases.org/


# Following lines from sections 3.3 through 3.5.4 of the text are needed as setup for Exercise 2
spamPath = "~/Stat480/RDataScience/SpamAssassinMessages"

dirNames = list.files(path = paste(spamPath, "messages", 
                                   sep = .Platform$file.sep))
fullDirNames = paste(spamPath, "messages", dirNames, 
                     sep = .Platform$file.sep)

indx = c(1:5, 15, 27, 68, 69, 329, 404, 427, 516, 852, 971)
fn = list.files(fullDirNames[1], full.names = TRUE)[indx]
sampleEmail = sapply(fn, readLines)

splitMessage = function(msg) {
  splitPoint = match("", msg)
  header = msg[1:(splitPoint-1)]
  body = msg[ -(1:splitPoint) ]
  return(list(header = header, body = body))
}

sampleSplit = lapply(sampleEmail, splitMessage)

getBoundary = function(header) {
  boundaryIdx = grep("boundary=", header)
  boundary = gsub('"', "", header[boundaryIdx])
  gsub(".*boundary= *([^;]*);?.*", "\\1", boundary)
}

dropAttach = function(body, boundary){
  
  bString = paste("--", boundary, sep = "")
  bStringLocs = which(bString == body)
  
  # if there are fewer than 2 beginning boundary strings, 
  # there is on attachment to drop
  if (length(bStringLocs) <= 1) return(body)
  
  # do ending string processing
  eString = paste("--", boundary, "--", sep = "")
  eStringLoc = which(eString == body)
  
  # if no ending boundary string, grab contents between the first 
  # two beginning boundary strings as the message body
  if (length(eStringLoc) == 0) 
    return(body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1)])
  
  # typical case of well-formed email with attachments
  # grab contents between first two beginning boundary strings and 
  # add lines after ending boundary string
  n = length(body)
  if (eStringLoc < n) 
    return( body[ c( (bStringLocs[1] + 1) : (bStringLocs[2] - 1), 
                     ( (eStringLoc + 1) : n )) ] )
  
  # fall through case
  # note that the result is the same as the 
  # length(eStringLoc) == 0 case, so code could be simplified by 
  # dropping that case and modifying the eStringLoc < n check to 
  # be 0 < eStringLoc < n
  return( body[ (bStringLocs[1] + 1) : (bStringLocs[2] - 1) ])
}

library(tm)
stopWords = stopwords()
cleanSW = tolower(gsub("[[:punct:]0-9[:blank:]]+", " ", stopWords))
SWords = unlist(strsplit(cleanSW, "[[:blank:]]+"))
SWords = SWords[ nchar(SWords) > 1 ]
stopWords = unique(SWords)

cleanText =
  function(msg)   {
    tolower(gsub("[[:punct:]0-9[:space:][:blank:]]+", " ", msg))
  }

findMsgWords = 
  function(msg, stopWords) {
    if(is.null(msg))
      return(character())
    
    words = unique(unlist(strsplit(cleanText(msg), "[[:blank:]\t]+")))
    
    # drop empty and 1 letter words
    words = words[ nchar(words) > 1]
    words = words[ !( words %in% stopWords) ]
    invisible(words)
  }

processAllWords = function(dirName, stopWords)
{
  # read all files in the directory
  fileNames = list.files(dirName, full.names = TRUE)
  # drop files that are not email, i.e., cmds
  notEmail = grep("cmds$", fileNames)
  if ( length(notEmail) > 0) fileNames = fileNames[ - notEmail ]
  
  messages = lapply(fileNames, readLines, encoding = "latin1")
  
  # split header and body
  emailSplit = lapply(messages, splitMessage)
  # put body and header in own lists
  bodyList = lapply(emailSplit, function(msg) msg$body)
  headerList = lapply(emailSplit, function(msg) msg$header)
  rm(emailSplit)
  
  # determine which messages have attachments
  hasAttach = sapply(headerList, function(header) {
    CTloc = grep("Content-Type", header)
    if (length(CTloc) == 0) return(0)
    multi = grep("multi", tolower(header[CTloc])) 
    if (length(multi) == 0) return(0)
    multi
  })
  
  hasAttach = which(hasAttach > 0)
  
  # find boundary strings for messages with attachments
  boundaries = sapply(headerList[hasAttach], getBoundary)
  
  # drop attachments from message body
  bodyList[hasAttach] = mapply(dropAttach, bodyList[hasAttach], 
                               boundaries, SIMPLIFY = FALSE)
  
  # extract words from body
  msgWordsList = lapply(bodyList, findMsgWords, stopWords)
  
  invisible(msgWordsList)
}

msgWordsList = lapply(fullDirNames, processAllWords, 
                      stopWords = stopWords) 
msgWordsList = unlist(msgWordsList, recursive = FALSE)


# Set working directory
setwd("~/Stat480/RDataScience/Chapter3")

# Load data structures and define variables and function needed in examples.
# Note that emailXX.rda is created in the section 3.8 code and spamAssassinDerivedDF.rda is created in the 
# section 3.9 setup code. If those files do not already exist, they will need to be created with the code from those
# respective sections.
# These are needed for Exercises 1, 3, and 4.
load("emailXX.rda")
indx = c(1:5, 15, 27, 68, 69, 329, 404, 427, 516, 852, 971)
sampleStruct = emailStruct[ indx ]
load("spamAssassinDerivedDF.rda")


##############################################################################################################


```

```{r,echo = FALSE}
fun = list(
  isSpam =
    expression(msg$isSpam)
  ,
  isRe =
    function(msg) {
      # Can have a Fwd: Re:  ... but we are not looking for this here.
      # We may want to look at In-Reply-To field.
      "Subject" %in% names(msg$header) && 
        length(grep("^[ \t]*Re:", msg$header[["Subject"]])) > 0
    }
  ,
  numLines =
    function(msg) length(msg$body)
  ,
  bodyCharCt =
    function(msg)
      sum(nchar(msg$body))
  ,
  underscore =
    function(msg) {
      if(!"Reply-To" %in% names(msg$header))
        return(FALSE)
      
      txt <- msg$header[["Reply-To"]]
      length(grep("_", txt)) > 0  && 
        length(grep("[0-9A-Za-z]+", txt)) > 0
    }
  ,
  subExcCt = 
    function(msg) {
      x = msg$header["Subject"]
      if(is.na(x))
        return(NA)
      
      sum(nchar(gsub("[^!]","", x)))
    }
  ,
  subQuesCt =
    function(msg) {
      x = msg$header["Subject"]
      if(is.na(x))
        return(NA)
      
      sum(nchar(gsub("[^?]","", x)))
    }
  ,
  numAtt = 
    function(msg) {
      if (is.null(msg$attach)) return(0)
      else nrow(msg$attach)
    }
  
  ,
  priority =
    function(msg) {
      ans <- FALSE
      # Look for names X-Priority, Priority, X-Msmail-Priority
      # Look for high any where in the value
      ind = grep("priority", tolower(names(msg$header)))
      if (length(ind) > 0)  {
        ans <- length(grep("high", tolower(msg$header[ind]))) >0
      }
      ans
    }
  ,
  numRec =
    function(msg) {
      # unique or not.
      els = getMessageRecipients(msg$header)
      
      if(length(els) == 0)
        return(NA)
      
      # Split each line by ","  and in each of these elements, look for
      # the @ sign. This handles
      tmp = sapply(strsplit(els, ","), function(x) grep("@", x))
      sum(sapply(tmp, length))
    }
  ,
  perCaps =
    function(msg)
    {
      body = paste(msg$body, collapse = "")
      
      # Return NA if the body of the message is "empty"
      if(length(body) == 0 || nchar(body) == 0) return(NA)
      
      # Eliminate non-alpha characters and empty lines 
      body = gsub("[^[:alpha:]]", "", body)
      els = unlist(strsplit(body, ""))
      ctCap = sum(els %in% LETTERS)
      100 * ctCap / length(els)
    }
  ,
  isInReplyTo =
    function(msg)
    {
      "In-Reply-To" %in% names(msg$header)
    }
  ,
  sortedRec =
    function(msg)
    {
      ids = getMessageRecipients(msg$header)
      all(sort(ids) == ids)
    }
  ,
  subPunc =
    function(msg)
    {
      if("Subject" %in% names(msg$header)) {
        el = gsub("['/.:@-]", "", msg$header["Subject"])
        length(grep("[A-Za-z][[:punct:]]+[A-Za-z]", el)) > 0
      }
      else
        FALSE
    },
  hour =
    function(msg)
    {
      date = msg$header["Date"]
      if ( is.null(date) ) return(NA)
      # Need to handle that there may be only one digit in the hour
      locate = regexpr("[0-2]?[0-9]:[0-5][0-9]:[0-5][0-9]", date)
      
      if (locate < 0)
        locate = regexpr("[0-2]?[0-9]:[0-5][0-9]", date)
      if (locate < 0) return(NA)
      
      hour = substring(date, locate, locate+1)
      hour = as.numeric(gsub(":", "", hour))
      
      locate = regexpr("PM", date)
      if (locate > 0) hour = hour + 12
      
      locate = regexpr("[+-][0-2][0-9]00", date)
      if (locate < 0) offset = 0
      else offset = as.numeric(substring(date, locate, locate + 2))
      (hour - offset) %% 24
    }
  ,
  multipartText =
    function(msg)
    {
      if (is.null(msg$attach)) return(FALSE)
      numAtt = nrow(msg$attach)
      
      types = 
        length(grep("(html|plain|text)", msg$attach$aType)) > (numAtt/2)
    }
  ,
  hasImages =
    function(msg)
    {
      if (is.null(msg$attach)) return(FALSE)
      
      length(grep("^ *image", tolower(msg$attach$aType))) > 0
    }
  ,
  isPGPsigned =
    function(msg)
    {
      if (is.null(msg$attach)) return(FALSE)
      
      length(grep("pgp", tolower(msg$attach$aType))) > 0
    },
  perHTML =
    function(msg)
    {
      if(! ("Content-Type" %in% names(msg$header))) return(0)
      
      el = tolower(msg$header["Content-Type"]) 
      if (length(grep("html", el)) == 0) return(0)
      
      els = gsub("[[:space:]]", "", msg$body)
      totchar = sum(nchar(els))
      totplain = sum(nchar(gsub("<[^<]+>", "", els )))
      100 * (totchar - totplain)/totchar
    },
  subSpamWords =
    function(msg)
    {
      if("Subject" %in% names(msg$header))
        length(grep(paste(SpamCheckWords, collapse = "|"), 
                    tolower(msg$header["Subject"]))) > 0
      else
        NA
    }
  ,
  subBlanks =
    function(msg)
    {
      if("Subject" %in% names(msg$header)) {
        x = msg$header["Subject"]
        # should we count blank subject line as 0 or 1 or NA?
        if (nchar(x) == 1) return(0)
        else 100 *(1 - (nchar(gsub("[[:blank:]]", "", x))/nchar(x)))
      } else NA
    }
  ,
  noHost =
    function(msg)
    {
      # Or use partial matching.
      idx = pmatch("Message-", names(msg$header))
      
      if(is.na(idx)) return(NA)
      
      tmp = msg$header[idx]
      return(length(grep(".*@[^[:space:]]+", tmp)) ==  0)
    }
  ,
  numEnd =
    function(msg)
    {
      # If we just do a grep("[0-9]@",  )
      # we get matches on messages that have a From something like
      # " \"marty66@aol.com\" <synjan@ecis.com>"
      # and the marty66 is the "user's name" not the login
      # So we can be more precise if we want.
      x = names(msg$header)
      if ( !( "From" %in% x) ) return(NA)
      login = gsub("^.*<", "", msg$header["From"])
      if ( is.null(login) ) 
        login = gsub("^.*<", "", msg$header["X-From"])
      if ( is.null(login) ) return(NA)
      login = strsplit(login, "@")[[1]][1]
      length(grep("[0-9]+$", login)) > 0
    },
  isYelling =
    function(msg)
    {
      if ( "Subject" %in% names(msg$header) ) {
        el = gsub("[^[:alpha:]]", "", msg$header["Subject"])
        if (nchar(el) > 0) nchar(gsub("[A-Z]", "", el)) < 1
        else FALSE
      }
      else
        NA
    },
  forwards =
    function(msg)
    {
      x = msg$body
      if(length(x) == 0 || sum(nchar(x)) == 0)
        return(NA)
      
      ans = length(grep("^[[:space:]]*>", x))
      100 * ans / length(x)
    },
  isOrigMsg =
    function(msg)
    {
      x = msg$body
      if(length(x) == 0) return(NA)
      
      length(grep("^[^[:alpha:]]*original[^[:alpha:]]+message[^[:alpha:]]*$", 
                  tolower(x) ) ) > 0
    },
  isDear =
    function(msg)
    {
      x = msg$body
      if(length(x) == 0) return(NA)
      
      length(grep("^[[:blank:]]*dear +(sir|madam)\\>", 
                  tolower(x))) > 0
    },
  isWrote =
    function(msg)
    {
      x = msg$body
      if(length(x) == 0) return(NA)
      
      length(grep("(wrote|schrieb|ecrit|escribe):", tolower(x) )) > 0
    },
  avgWordLen =
    function(msg)
    {
      txt = paste(msg$body, collapse = " ")
      if(length(txt) == 0 || sum(nchar(txt)) == 0) return(0)
      
      txt = gsub("[^[:alpha:]]", " ", txt)
      words = unlist(strsplit(txt, "[[:blank:]]+"))
      wordLens = nchar(words)
      mean(wordLens[ wordLens > 0 ])
    }
  ,
  numDlr =
    function(msg)
    {
      x = paste(msg$body, collapse = "")
      if(length(x) == 0 || sum(nchar(x)) == 0)
        return(NA)
      
      nchar(gsub("[^$]","", x))
    }
)
SpamCheckWords =
  c("viagra", "pounds", "free", "weight", "guarantee", "million", 
    "dollars", "credit", "risk", "prescription", "generic", "drug",
    "financial", "save", "dollar", "erotic", "million", "barrister",
    "beneficiary", "easy", 
    "money back", "money", "credit card")


getMessageRecipients =
  function(header)
  {
    c(if("To" %in% names(header))  header[["To"]] else character(0),
      if("Cc" %in% names(header))  header[["Cc"]] else character(0),
      if("Bcc" %in% names(header)) header[["Bcc"]] else character(0)
    )
  }
```

#Exercise 1
```{r}
funcList = c(fun,VowelPerc =
  function(msg)
  {
    txt = paste(msg$body, collapse = "")
    txt = gsub("[^[:alpha:]]","",txt)
    txt = tolower(txt)
    txt = strsplit(txt,'')
    if(length(txt) == 0 || sum(nchar(txt)) == 0) {return(NA)}
    words = unlist(txt)
    n = 0
    for (i in words){
      if (i == 'a' || i == 'e' || i == 'i' || i == 'o' || i == 'u'){
        n = n + 1
      }
    }
    wordLens = sum(nchar(words))
    (n/wordLens)*100
  })

createDerivedDF =
  function(email = emailStruct, operations = funcList, 
           verbose = FALSE)
  {
    els = lapply(names(operations),
                 function(id) {
                   if(verbose) print(id)
                   e = operations[[id]]
                   v = if(is.function(e)) 
                     sapply(email, e)
                   else 
                     sapply(email, function(msg) eval(e))
                   v
                 })
    
    df = as.data.frame(els)
    names(df) = names(operations)
    invisible(df)
  }
emailDF = createDerivedDF(emailStruct)

sampleDF = createDerivedDF(sampleStruct)
sampleDF$VowelPerc
emailDF$VowelPerc[is.na(emailDF$VowelPerc)==TRUE]  # Check NA existance; None
emailDF$VowelPerc[emailDF$VowelPerc>100 || emailDF$VowelPerc<0] #Check unreasonable value; None

spam = subset(emailDF, isSpam == TRUE)
ham = subset(emailDF, isSpam == FALSE)

(spam_min=min(spam$VowelPerc)) #Min percentage for Spam 6.667%
(spam_max=max(spam$VowelPerc)) #Max percentage for Spam 45.54%
(ham_min=min(ham$VowelPerc)) #Min percentage for Ham 19.28%
(ham_max=max(ham$VowelPerc)) #Max percentage for Ham 43.06%

boxplot(ham$VowelPerc,spam$VowelPerc, main = 'Percent of Vowels Letter Comparison of Ham and Spam',
        names=c('Ham','Spam'), ylab = 'Percent of Vowels')
```

From the boxplot shown above, the difference between ham and spam is not significant, either based on the mean difference or the range difference. Also, there is no significantly different in space between these two boxplots, and big portion of these two boxplots is overlapping. Thus, I believe that it is not a useful feature to classify ham and spam. 


#Exercise 2

```{r}
avgLength=function(msgWords){
  avgs=mean(sapply(msgWords,function(msgWord){nchar(msgWord, keepNA=TRUE)}))
}
avg=sapply(msgWordsList, avgLength)
mean(avg)
range(avg)

wordLen = emailDF$avgWordLen
mean(wordLen)
range(wordLen)
```

Firstly, the technique with findMsgWords use the stopwords, such as or, yes, which, and the findMsgWords will remove those words from the text, which has the potential influence to higher the average of the length (5.835972 > 4.488554). 

Plus, the algorithm of findMsgWords use the unique() function, so it gets rid of the repetitive words, so there might be some chance to affect the average words length. 

Also the range of both part is different as shown above, which (2.82, 26.0) for findMsgWords and avgLength: (1.36, 26.0) 

Overall, based on the analysis, we are unlikely use findMsgWords function to find the average length.

#Exercise 3

```{r}
set.seed(418910)
library(rpart)
setupRpart = function(data) {
  logicalVars = which(sapply(data, is.logical))
  facVars = lapply(data[ , logicalVars], 
                   function(x) {
                     x = as.factor(x)
                     levels(x) = c("F", "T")
                     x
                   })
  cbind(facVars, data[ , - logicalVars])
}

# Process the email data frame.
emailDFrp = setupRpart(emailDF)

numSpam = sum(emailDFrp$isSpam=='T')
numHam = sum(emailDFrp$isSpam=='F')
# Obtain the Type I and Type II error rates from these fittings.

testSpamIdx = sample(numSpam, size = floor(numSpam/3))
testHamIdx = sample(numHam, size = floor(numHam/3))

testDF = 
  rbind( emailDFrp[ emailDFrp$isSpam == "T", ][testSpamIdx, ],
         emailDFrp[emailDFrp$isSpam == "F", ][testHamIdx, ] )
trainDF =
  rbind( emailDFrp[emailDFrp$isSpam == "T", ][-testSpamIdx, ], 
         emailDFrp[emailDFrp$isSpam == "F", ][-testHamIdx, ])


# Fit the recursive partitioning model for spam as a function of all variables in the data frame.
rpartFit = rpart(isSpam ~ ., data = trainDF, method = "class")

complexityVals = c(seq(1,100,1))
fits = lapply(complexityVals, function(x) {
  rpartObj = rpart(isSpam ~ . -VowelPerc , data = trainDF,
                   method="class", 
                   control = rpart.control(minsplit = 1, minbucket = x ) )
  
  predict(rpartObj, 
          newdata = testDF[ , names(testDF) != "isSpam"],
          type = "class")
})

spam = testDF$isSpam == "T"
numSpam = sum(spam)
numHam = sum(!spam)
errs = sapply(fits, function(preds) {
  typeI = sum(preds[ !spam ] == "T") / numHam
  typeII = sum(preds[ spam ] == "F") / numSpam
  c(typeI = typeI, typeII = typeII)
})

library(RColorBrewer)
cols = brewer.pal(9, "Set1")[c(3, 4, 5)]
plot(errs[1,] ~ complexityVals, type="l", col=cols[2], 
     lwd = 2, ylim = c(0,0.25),
     ylab="Error", xlab="Minbucket Values", main = "Error Rate vs Minbucket")
points(errs[2,] ~ complexityVals, type="l", col=cols[1], lwd = 2)


text(x =c(15, 15.1), y = c(0.12, 0.03), 
     labels=c("Type II Error", "Type I Error"))

minI = which(errs[1,] == min(errs[1,]))[1]
abline(v = complexityVals[minI], col ="grey", lty =3, lwd=2)

text(3, errs[1, minI]+0.02, 
     formatC(errs[1, minI], digits = 2))
text(3, errs[2, minI]+0.02, 
     formatC(errs[2, minI], digits = 3))

which(errs[1,] == min(errs[1,])) #Minbucket value that has the minimum error rate.
```

The trend of the plot is that the error for both Type 1 and type 2 increased along with the increasing of minbucket value except the first 31 values. As we could see from the plot and the function returns, with the value from 1 to 31 for minbucket, the error rate are the same, which is 0.054. However, if minbucket is 1, there might overfitting happening in this case, thus, we perfer the relatively safe approach, which we would choose 31 as the approprociate minbucket value. 

